\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{conjecture}{Conjecture}
\newtheorem{hypothesis}{Hypothesis}
\newtheorem{observation}{Observation}
\newtheorem{justification}{Justification}
\newtheorem{designprinciple}{Design Principle}
\newtheorem{example}{Example}

\title{OmegaSim/OmegaHive Strange-Attractor Tuning:\\
A Hyperseed-Dynamical Formalization}
\author{OmegaClaw / ZeroBot working notes for Ben Goertzel}
\date{2026-06-28}

\begin{document}
\maketitle

\begin{abstract}
This note formalizes a useful mathematical frame for constructing and tuning
OmegaSim/OmegaHive-style simulations so that they can explore complex bounded
nonlinear dynamics, including operational strange-attractor regimes.  The core
idea is to model hives as delayed, coupled, dissipative nonlinear systems; to
express tuning through dimensionless control ratios; to prove what follows from
well-posedness, boundedness, contraction, and delay-instability arguments; and to
translate the resulting theory into concrete simulation recommendations.  The
claims here are deliberately split into theorems, conjectures, operational
criteria, and implementation suggestions.
\end{abstract}

\section{Provenance and aim}

This note responds to Ben Goertzel's 2026-06-28 request for a rigorous
mathematical formalization of the ``useful frame'' proposed for tuning one-hive
and OmegaHive-style simulations toward complex strange-attractor dynamics.  The
requested steps are:
\begin{enumerate}
  \item formalize concepts bridging Hyperseed and nonlinear dynamics;
  \item state and prove general-purpose theorems applicable across agent hives;
  \item specialize these results to OmegaHive1/OmegaSim-style simulations;
  \item give theory-motivated instructions for the Codex/OmegaSim simulation
        agent.
\end{enumerate}

The purpose is not to prove that a given current simulation has a strange
attractor.  Rather, it is to identify mathematically meaningful design knobs and
verification gates so that such a claim can later become evidence-bearing.

\section{S1. Concepts bridging Hyperseed and nonlinear dynamics}

\subsection{Hyperseed event schema as a state observable}

\begin{definition}[Hyperseed process event]
A Hyperseed-relevant process event is a typed record
\[
  e = \langle \iota, \tau, A, C, R, P, \Theta, \nu \rangle
\]
where $\iota$ is identity, $\tau$ time, $A$ the active agents or sub-processes,
$C$ context, $R$ realized relation or result, $P$ provenance, $\Theta$ an
interpretive type assignment, and $\nu$ uncertainty/truth-value metadata.
\end{definition}

\begin{definition}[Hive observable]
A hive observable is a map
\[
  O : E_{\le t} \to X \subseteq \mathbb R^n
\]
from a finite or filtered history of Hyperseed process events to a numerical
state vector.  Components may include agent activation, resource load, attention,
coordination phase, memory pressure, trust, artifact quality, task backlog,
message volume, or semantic diversity.
\end{definition}

\begin{definition}[Scrutable reduction]
A reduction from Hyperseed events to dynamical state is scrutable when each
coordinate $x_i$ of $x=O(E)$ has: (i) a declared meaning, (ii) a measurement rule,
(iii) provenance to events or logs, and (iv) an uncertainty model.  Unscrutable
coordinates may be useful for machine learning, but should not carry ontological
interpretation without an additional explanation layer.
\end{definition}

\subsection{Delayed agent-hive dynamics}

Let $x(t)\in X\subseteq\mathbb R^n$ be a numerical hive state derived from the
observable map above.  Let the delayed history segment be
\[
  x_t(\theta)=x(t+\theta),\qquad \theta\in[-\tau,0],
\]
where $\tau\ge0$ is a maximum communication, reaction, or memory delay.

\begin{definition}[Continuous delayed hive system]
A continuous delayed hive system is a retarded functional differential equation
\[
  \dot x(t)=F(x_t,u(t),\eta(t);\theta),
\]
where $F:C([ -\tau,0],X)\times U\times E\to\mathbb R^n$, $u(t)$ is a controllable
intervention/tuning signal, $\eta(t)$ is bounded environmental or stochastic
forcing, and $\theta$ denotes fixed parameters.
\end{definition}

\begin{definition}[Discrete delayed hive simulation]
A discrete delayed hive simulation is a lifted map
\[
  x_{k+1}=G(x_k,x_{k-1},\ldots,x_{k-m},u_k,\eta_k;\theta),
\]
with lifted state
\[
  z_k=(x_k,x_{k-1},\ldots,x_{k-m})\in X^{m+1}.
\]
Equivalently,
\[
  z_{k+1}=\widehat G(z_k,u_k,\eta_k;\theta)
  =(G(x_k,\ldots,x_{k-m},u_k,\eta_k),x_k,\ldots,x_{k-m+1}).
\]
\end{definition}

\subsection{Dimensionless tuning coordinates}

The raw constants of a hive simulation are usually too numerous to sweep
intelligently.  We therefore define dimensionless ratios.

\begin{definition}[Dimensionless hive control ratios]
For a delayed nonlinear hive model, define:
\[
  \rho = g\,r(A),
  \qquad
  \delta = \frac{\tau}{T_{\rm relax}},
  \qquad
  \mu = \frac{T_{\rm memory}}{T_{\rm decision}},
\]
\[
  \kappa = \frac{\|A_{\rm cross}\|}{\|A_{\rm within}\|},
  \qquad
  \nu = \frac{\sigma_\eta}{\|F_0\|+\epsilon},
\]
where $g$ is a typical nonlinear response gain, $r(A)$ is the spectral radius of
an interaction matrix or linearized coupling operator, $T_{\rm relax}$ is local
relaxation time, $T_{\rm memory}$ memory decay time, $T_{\rm decision}$ update
time, $A_{\rm cross}$ and $A_{\rm within}$ are cross-hive and within-hive
couplings, and $\sigma_\eta$ is noise magnitude.
\end{definition}

Interpretation: $\rho$ measures feedback amplification, $\delta$ delay pressure,
$\mu$ persistence, $\kappa$ modular leakage, and $\nu$ noise-to-drift.  These are
not universal constants; they are organizing coordinates for phase diagrams.

\begin{definition}[Operational strange-attractor regime]
A parameter region $\mathcal P$ is an operational strange-attractor regime for a
simulation if, for a specified observable $O$ and lifted state $z$, the following
hold robustly over validation runs:
\begin{enumerate}
  \item trajectories are ultimately bounded and not merely saturating at hard
        numerical limits;
  \item the largest finite-time Lyapunov estimate is positive under controlled
        perturbations of initial histories;
  \item long-run statistics are stable under timestep/refinement and seed checks;
  \item recurrence plots, broadband spectra, or reconstructed dimensions indicate
        non-periodic structured recurrence rather than white noise.
\end{enumerate}
This is an empirical criterion.  It is weaker than a proof of a mathematical
strange attractor.
\end{definition}

\section{S2. General theorems}

\subsection{Well-posedness and boundedness}

\begin{theorem}[Local well-posedness for delayed hives]
Assume $F$ is locally Lipschitz in the history variable $x_t$, uniformly over
bounded $u$ and $\eta$.  Then for every continuous initial history
$\phi\in C([ -\tau,0],X)$ there exists a unique maximal solution.
\end{theorem}

\begin{proof}
This is the Picard--Lindelof theorem for retarded functional differential
equations.  The history space $C([ -\tau,0],X)$ is a Banach space under the sup
norm.  Local Lipschitz continuity gives a contraction on a sufficiently short
interval.  Repetition extends the solution until it exits every bounded subset of
the state space.
\end{proof}

\begin{theorem}[Ultimate boundedness from dissipativity]
Suppose there is a radially unbounded $C^1$ function
$V:\mathbb R^n\to\mathbb R_{\ge0}$ and constants $a,b>0$ such that along all
trajectories with bounded input magnitude $M$,
\[
  \dot V(x(t))\le -aV(x(t))+bM.
\]
Then
\[
  \limsup_{t\to\infty}V(x(t))\le \frac{bM}{a}.
\]
\end{theorem}

\begin{proof}
By Gronwall's inequality,
\[
  V(t)\le e^{-at}V(0)+\frac{bM}{a}(1-e^{-at}).
\]
Taking $t\to\infty$ gives the result.
\end{proof}

\begin{corollary}[Attractor-talk requires a bounded regime]
A simulation run that diverges, explodes numerically, or only bounces against
uninterpreted clipping bounds does not by itself justify talk of an attractor.
Dissipativity or empirically verified boundedness is a prerequisite diagnostic.
\end{corollary}

\begin{theorem}[Forward-invariant box for delayed maps]
For
\[
  x_{k+1}=G(x_k,\ldots,x_{k-m},u_k),
\]
assume there is a compact set $K\subseteq\mathbb R^n$ such that
\[
  G(K^{m+1},U)\subseteq K.
\]
Then $K^{m+1}$ is forward invariant under the lifted map $\widehat G$.
\end{theorem}

\begin{proof}
If $z_k=(x_k,\ldots,x_{k-m})\in K^{m+1}$, then each delayed component lies in $K$.
By assumption the new component $x_{k+1}=G(\cdots)$ lies in $K$.  The remaining
components of $z_{k+1}$ are shifted copies of previous components in $K$.
Therefore $z_{k+1}\in K^{m+1}$.
\end{proof}

\subsection{Contraction: when complex dynamics is impossible}

\begin{theorem}[Delayed-map contraction implies a unique attracting history]
Let $\widehat G:K^{m+1}\to K^{m+1}$ be the lifted delayed map on compact convex
$K^{m+1}$, and suppose it is a contraction with Lipschitz constant $L<1$ under
some norm.  Then there is a unique fixed lifted state $z^\ast$, and every
trajectory converges to $z^\ast$.
\end{theorem}

\begin{proof}
This is Banach's fixed-point theorem applied to $\widehat G$ on the complete
metric space $K^{m+1}$.  Iterates of a contraction converge to the unique fixed
point.
\end{proof}

\begin{corollary}[Low-gain hives cannot be chaotic]
If the effective lifted update of a hive simulation is contractive, then it cannot
have a strange attractor, quasiperiodic orbit, or sustained multi-cycle behavior.
Therefore one route toward complex dynamics is to increase gain, delay,
cross-coupling, or memory persistence until the lifted system leaves the
contraction regime while preserving boundedness.
\end{corollary}

\begin{proposition}[A practical Lipschitz bound]
Suppose
\[
  G(x_k,\ldots,x_{k-m})=S\left(B+\sum_{j=0}^m A_j x_{k-j}\right)
\]
where $S$ is componentwise Lipschitz with constant $s$ and the lifted norm is
chosen so that the shift has unit weight.  A sufficient condition for contraction
is that the induced lifted Lipschitz constant of the companion map is less than
one; in particular the response gain $s$ and delayed coupling norms $\|A_j\|$
provide an upper bound.  Thus the product ``gain times coupling'' is the first
order control axis.
\end{proposition}

\begin{proof}
The non-shift component changes by at most
$s\sum_j\|A_j\|\,\|\Delta x_{k-j}\|$ under the chosen induced norm, while the shift
components copy previous differences.  Weighted norms can penalize older delay
coordinates so that the whole companion map is contractive when the response and
coupling norms are sufficiently small.  This proves the sufficient, not necessary,
condition.
\end{proof}

\subsection{Delay-induced oscillatory instability}

\begin{theorem}[Scalar delay Hopf boundary]
Consider
\[
  \dot x(t)=-a x(t)+b x(t-\tau),\qquad a>0.
\]
The characteristic equation is
\[
  \lambda+a-b e^{-\lambda\tau}=0.
\]
A Hopf crossing candidate satisfies $\lambda=i\omega$, hence
\[
  a=b\cos(\omega\tau),\qquad
  \omega=-b\sin(\omega\tau).
\]
Therefore delayed feedback can destabilize an equilibrium through an oscillatory
root crossing even when the local undelayed term is dissipative.
\end{theorem}

\begin{proof}
Substitute $x(t)=e^{\lambda t}$ to obtain the characteristic equation.  At an
oscillatory stability boundary $\lambda=i\omega$.  Separating real and imaginary
parts gives the two displayed equations.
\end{proof}

\begin{corollary}[Why $\delta=\tau/T_{\rm relax}$ matters]
The ratio of communication/memory delay to local relaxation time is a genuine
bifurcation coordinate, not just an implementation detail.  Increasing delay can
convert damped consensus into oscillation and, in higher-dimensional nonlinear
systems, into period-doubling or more complex regimes.
\end{corollary}

\subsection{Existence of compact attractors}

\begin{theorem}[Compact global attractor under dissipative smoothing]
For a continuous delayed hive semiflow on $C([ -\tau,0],X)$, assume: (i) unique
forward solutions exist, (ii) solutions enter a bounded absorbing set, and (iii)
the semiflow is asymptotically compact.  Then there is a compact global attractor
$\mathcal A$ attracting all bounded sets.
\end{theorem}

\begin{proof}
This is the standard global-attractor theorem for continuous semiflows: a bounded
absorbing set plus asymptotic compactness implies the omega-limit of the absorbing
set is nonempty, compact, invariant, and attracts every bounded set.
\end{proof}

\begin{corollary}[Delay makes rigor harder]
Delayed systems are infinite-dimensional in history space.  Compactness and low
fractal dimension are not automatic.  OmegaSim should therefore treat finite
state-history simulations as approximations and validate attractor claims under
history-length and timestep variation.
\end{corollary}

\subsection{What remains conjectural}

\begin{conjecture}[Useful complex regime window]
For many OmegaHive-like systems there is a practical window between contraction
and blow-up/saturation, characterized roughly by $\rho\approx1$, non-negligible
$\delta$, moderate $\mu$, and controlled $\kappa$, in which bounded complex
recurrence is most likely.
\end{conjecture}

This is not a theorem because it depends on nonlinear form, topology, stochastic
forcing, and measurement.  It is a theory-guided search heuristic.

\begin{conjecture}[Hyperseed-scrutable complexity]
The most scientifically useful OmegaSim regimes will be those where the
low-dimensional observables show complex recurrence while Hyperseed event records
remain interpretable enough to explain which agent roles, memories, artifacts,
and feedback loops generate the recurrence.
\end{conjecture}

\section{S3. Specialization to OmegaHive1 / OmegaSim}

Since the exact OmegaHive1 simulation state is still evolving, define a generic
OmegaHive1 state vector as
\[
  x_k=(a_k,r_k,m_k,q_k,h_k,c_k,p_k)\in\mathbb R^n,
\]
where $a$ is agent activation/attention, $r$ resource or compute load, $m$ memory
persistence/pressure, $q$ artifact quality or progress, $h$ semantic diversity,
$c$ coordination/coherence, and $p$ task pressure.  The components may be vectors
indexed by agents, roles, artifacts, or sub-hives.

A useful discrete OmegaHive1 skeleton is
\[
  x_{k+1}=S\left(W_0x_k+W_1x_{k-d}+W_m M_k+B u_k+\xi_k\right)-D x_k,
\]
with memory update
\[
  M_{k+1}=(1-\alpha)M_k+\alpha H(x_k,\text{events}_k),
\]
where $S$ is saturating nonlinear response, $W_0$ present-time coupling, $W_1$
delayed coupling, $M_k$ memory state, $u_k$ interventions, $\xi_k$ controlled
noise, and $D$ local decay.

\begin{proposition}[OmegaHive1 boundedness by saturation and resource accounting]
If each component of $S$ maps into a compact interval $K_i$ and resource/memory
updates are either projected into compact budgets or dissipative in the sense of
the boundedness theorem, then the lifted OmegaHive1 history has a compact forward
invariant set.
\end{proposition}

\begin{proof}
The saturating components satisfy the invariant-box theorem.  Resource and memory
coordinates either remain in explicit compact budgets or satisfy dissipative
bounds.  The product of these compact coordinate sets is compact and invariant
under the lifted delayed map.
\end{proof}

\begin{proposition}[OmegaHive1 contraction baseline]
If the fitted or designed Lipschitz constant of the lifted OmegaHive1 update is
below one, the system will converge to a unique stable regime and should not be
expected to exhibit strange-attractor-like dynamics.
\end{proposition}

\begin{proof}
Immediate from the delayed-map contraction theorem.
\end{proof}

\begin{proposition}[OmegaHive1 delay/gain destabilization target]
Let $x^\ast$ be an equilibrium of a smooth deterministic OmegaHive1 skeleton.
Linearizing gives
\[
  \delta x_{k+1}=J_0\delta x_k+J_1\delta x_{k-d}+J_m\delta M_k.
\]
The stability of $x^\ast$ is determined by the eigenvalues of the lifted companion
Jacobian.  Parameter sweeps that increase response gain, delayed coupling, or
memory persistence move these eigenvalues toward or beyond the unit circle.
\end{proposition}

\begin{proof}
Linearization of the lifted map gives an ordinary finite-dimensional map on
history and memory coordinates.  Local discrete-time stability is equivalent to
all eigenvalues of the companion Jacobian lying inside the unit disk.  Increasing
the norms of coupling and persistence terms changes this Jacobian continuously;
instability occurs when eigenvalues cross the unit circle.
\end{proof}

\subsection{OmegaHive1 order parameters}

For diagnosis, define low-dimensional order parameters:
\[
  R_k=\left|\frac{1}{N}\sum_{j=1}^N e^{i\phi_{j,k}}\right|
  \quad\text{(coordination/synchrony)},
\]
\[
  H_k=-\sum_a p_k(a)\log p_k(a)
  \quad\text{(behavioral or semantic diversity)},
\]
\[
  C_k=x_k^\top L x_k
  \quad\text{(network tension across graph Laplacian $L$)},
\]
\[
  Q_k=\text{artifact-progress or quality metric},
  \qquad
  B_k=\text{resource/backlog pressure}.
\]
Analyze trajectories in $(R,H,C,Q,B)$ as well as in raw state space.

\begin{designprinciple}[Do not search for chaos in raw logs first]
Raw agent traces are too high-dimensional and semantically mixed.  First define
scrutable observables and order parameters, then test whether the induced
observable dynamics has bounded recurrence, sensitivity, and nonperiodicity.
\end{designprinciple}

\section{S4. Theory-motivated OmegaSim instructions}

\subsection{Implementation targets}

\begin{enumerate}
  \item Implement the lifted delayed state $z_k=(x_k,\ldots,x_{k-m},M_k)$
        explicitly.  Do not hide delays in ad hoc buffers without logging them.
  \item Define a compact state budget.  Use saturating nonlinearities,
        normalization, or explicit resource conservation so runs cannot be
        mistaken for attractors while diverging.
  \item Expose the control ratios $\rho,\delta,\mu,\kappa,\nu$ as primary sweep
        parameters.  Raw constants should be derived from these ratios where
        possible.
  \item Compute the fixed point or empirical baseline regime at low gain.  Use it
        as the contraction/control case.
  \item Linearize the deterministic skeleton around equilibria or recurrent
        regimes.  Track eigenvalues of the lifted companion Jacobian as parameters
        change.
  \item Run sweeps that cross from $\rho<1$ to $\rho>1$, from $\delta\approx0$ to
        $\delta=O(1)$, and from weak to moderate memory persistence $\mu$.
  \item Near transitions, run hysteresis sweeps upward and downward in gain/delay
        to detect multistability and crises.
  \item Estimate finite-time Lyapunov exponents with paired trajectories, periodic
        renormalization, multiple perturbation directions, and matched noise
        paths.
  \item Add null controls: no-delay, linearized nonlinearity, degree-preserving
        rewiring, delay-shuffled, phase-randomized surrogate, and frozen-noise
        paired comparisons.
  \item Classify each region as fixed point, limit cycle, quasiperiodic,
        candidate chaotic, transient chaotic, noise-driven irregular, divergent,
        or trivial saturation.
\end{enumerate}

\subsection{Specific suggested sweep grid}

A first OmegaSim phase diagram should sweep:
\[
  \rho\in[0.2,3.0],\qquad
  \delta\in\{0,0.1,0.3,1,3\},
\]
\[
  \mu\in\{0.5,1,3,10\},\qquad
  \kappa\in\{0,0.05,0.2,0.8\},
  \qquad
  \nu\in\{0,10^{-3},10^{-2},10^{-1}\}.
\]
The exact ranges should be rescaled after measuring the relaxation time and
state norms of the implemented simulator.

\subsection{Acceptance criteria for a candidate strange-attractor region}

A region should only be reported as a candidate complex/strange-attractor regime
if it passes all of these gates:
\begin{enumerate}
  \item boundedness gate: norms and resource variables remain in a nontrivial
        compact range;
  \item refinement gate: the qualitative regime survives smaller timestep or
        longer history resolution;
  \item perturbation gate: positive finite-time Lyapunov estimate under fixed
        noise path;
  \item surrogate gate: behavior differs from no-delay, linearized, and
        phase-randomized controls;
  \item semantic gate: Hyperseed event records identify plausible feedback loops
        generating the dynamics;
  \item task gate: the regime is related to useful simulation behavior, not just
        aesthetic irregularity.
\end{enumerate}

\section{Conclusion}

The rigorous core is modest but useful.  If the hive update is ill-posed,
unbounded, or contractive, complex attractor claims are unsupported or impossible.
If the system is well-posed, bounded, non-contractive, delayed, and nonlinear,
then gain, delay, memory persistence, modular leakage, and noise become principled
coordinates for exploring bifurcations and candidate strange-attractor regimes.

The practical OmegaSim instruction is therefore: build the simulator so these
coordinates are explicit; enforce boundedness; sweep through the contraction
boundary; test delay-induced instability; and only call a regime
strange-attractor-like after Lyapunov, recurrence, surrogate, refinement, and
semantic-provenance checks pass.

\input{omegasim_feedback_revision_2026_07_01}

\end{document}
