\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[hidelinks]{hyperref}

\newtheorem{definition}{Definition}
\newtheorem{claim}{Claim}
\newtheorem{conjecture}{Conjecture}
\newtheorem{theorem}{Theorem}

\title{Hyperseed Formalization Draft: ``Tag, You're Not It''}
\author{Source: Ben Goertzel; formalization draft by ProtomegaTron/ZeroBot}
\date{Source published 2026-06-26; draft retrieved/formalized 2026-06-28}

\begin{document}
\maketitle

\section{Source and status}
Source article: \href{https://bengoertzel.substack.com/p/tag-youre-not-it}{``Tag, You're Not It''}, Eurykosmotron, by Ben Goertzel.
This is an initial Hyperseed-oriented formalization. It paraphrases the article's main intellectual moves and separates source claims from proposed formal objects.

\section{Informal Hyperseed reading}
The article contrasts two patterns for persistent AI teammates. Claude Tag is treated as a polished, Slack-bound role-agent: a channel wears a managed Claude identity, backed by service accounts, billing, audit logs, and ambient/asynchronous tasking. OmegaClaw/OmegaHive are framed as agent-first societies: named cognitive principals with explicit memories, roles, tool boundaries, and inspectable inter-agent process.

The Hyperseed-relevant axis is therefore not merely ``which product is better,'' but how cognitive work becomes embodied in a substrate. A platform-bound agent makes the communication surface the organizing ontology. An agent-first society makes the agent, its identity, its memory, and its relations to other agents the organizing ontology.

\section{Core formal objects}

\begin{definition}[Agentic work substrate]
An agentic work substrate is a tuple
\[
S = \langle A, C, M, T, P, O, B \rangle
\]
where $A$ is a set of agents or agent-roles, $C$ communication contexts, $M$ memory structures, $T$ tools/actions, $P$ permission and policy structures, $O$ observation/audit structures, and $B$ billing/resource constraints.
\end{definition}

\begin{definition}[Platform-bound role-agent]
A platform-bound role-agent is an agentic work substrate in which a platform context $c \in C$ determines the agent's identity, memory boundary, and operational scope. Claude Tag, as described in the article, is modeled as approximately platform-bound because Slack channels provide the key compartmentalizing structure.
\end{definition}

\begin{definition}[Agent-first cognitive society]
An agent-first cognitive society is an agentic work substrate in which persistent named agents $a \in A$ are primary principals, each with role, memory, permissions, and remit, and communication surfaces are secondary transports. OmegaClaw/OmegaHive are described as aiming at this pattern.
\end{definition}

\begin{definition}[Inspectable durable memory]
A memory structure $M$ is inspectably durable when its stored contents, provenance, update history, and influence on current action can be queried or reconstructed by collaborators and verifier processes. Hyperon AtomSpace-style symbolic metagraph memory is proposed as a candidate implementation family.
\end{definition}

\begin{definition}[Live verifier]
A live verifier is an observer process $v$ with read-only or policy-scoped access to agent reasoning traces, actions, permissions, and provenance, such that oversight occurs during process execution rather than only after completion via logs.
\end{definition}

\section{Claims as formal contrasts}

\begin{claim}[Wrapper novelty versus substrate novelty]
The main novelty of Claude Tag is wrapper-level product integration rather than a fundamentally new agentic-work substrate.
\end{claim}

\begin{claim}[Opaque memory risk]
If an autonomous long-running agent's memory is opaque to users, then the agentic work substrate has elevated risk of hidden bias, sycophancy, stale context, and unexplained behavior.
\end{claim}

\begin{claim}[Attribution requirement]
For audit-grade multi-agent work, agent identities must remain distinguishable across external integrations. A collapse from multiple contextual agents to a single upstream app identity weakens attribution.
\end{claim}

\begin{claim}[Open ecosystem advantage]
An open, model-agnostic, transport-agnostic ecosystem has more degrees of freedom for solving durable memory, safe runtime isolation, attribution, and oversight than a closed platform-bound product.
\end{claim}

\begin{claim}[OmegaClaw design direction]
OmegaClaw is characterized by named persistent principals with symbolic memory and cognition layers; OmegaHive extends this into a society of cognitive and service agents coordinated by explicit task, permission, promotion, and inspection structures.
\end{claim}

\section{Candidate Hyperseed schema}
Let $\mathsf{Embodies}(x,y)$ mean that system $x$ embodies pattern $y$; $\mathsf{Requires}(x,y)$ mean that reliable realization of $x$ requires property $y$; and $\mathsf{Supports}(x,y)$ mean that mechanism $x$ supports property $y$.

\[
\begin{aligned}
&\mathsf{Embodies}(\mathsf{ClaudeTag}, \mathsf{PlatformBoundRoleAgent}) \\
&\mathsf{Embodies}(\mathsf{OmegaClaw}, \mathsf{AgentFirstCognitiveSociety}) \\
&\mathsf{Requires}(\mathsf{TrustworthyAgenticWork}, \mathsf{InspectableDurableMemory}) \\
&\mathsf{Requires}(\mathsf{TrustworthyAgenticWork}, \mathsf{AuditGradeAttribution}) \\
&\mathsf{Requires}(\mathsf{TrustworthyAgenticWork}, \mathsf{LiveProtocolOversight}) \\
&\mathsf{Supports}(\mathsf{HyperonAtomSpaceMemory}, \mathsf{InspectableDurableMemory}) \\
&\mathsf{Supports}(\mathsf{LiveVerifier}, \mathsf{LiveProtocolOversight}).
\end{aligned}
\]

\section{Claim-indexed Hyperseed formalizations}
The labels C1--C17 correspond to paragraphs 1--17 of the claim inventory in
\texttt{claims.md}, each paraphrasing the cited article.  ``Definition'' marks
an introduced formal object, ``Claim'' an article-triggered proposition,
``Theorem'' a consequence of the stated model, and ``Conjecture'' a proposed
but unproved design hypothesis.

\begin{definition}[C1: Persistent teammate pattern]
Let $\mathcal{E}$ be an agent ecosystem and let
$\Pi(a)=\langle\iota_a,\mu_a,\chi_a,\tau_a\rangle$ denote agent $a$'s
persistent identity, memory, communication contexts, and tools.  A
\emph{persistent teammate pattern} exists when $\Pi(a)$ remains addressable
across a nontrivial sequence of tasks.  Claude Tag is an instance of this
previously explored open-ecosystem pattern, with comparatively polished
packaging.  \emph{Provenance: article; claim-inventory paragraph 1.}
\end{definition}

\begin{definition}[C2: Wrapper contribution]
For a product $x$, decompose its capabilities as
$K(x)=K_{\mathrm{sub}}(x)\cup K_{\mathrm{wrap}}(x)$, where substrate
capabilities implement cognition and persistence, while wrapper capabilities
implement managed identity, transport integration, billing, asynchronous
dispatch, and operational polish.  The article assesses Tag's distinctive
contribution as concentrated in $K_{\mathrm{wrap}}$.
\emph{Provenance: article; claim-inventory paragraph 2.}
\end{definition}

\begin{definition}[C3: Shared proactive agent]
An agent $a$ is shared and proactive when it has durable memory $\mu_a$,
initiative policy $I_a$, tools $T_a$, a shared audience $U_a$, and a stable
address $n_a$, so that users may delegate by $n_a$ and $a$ may initiate
policy-permitted actions.  This, rather than any particular chat transport,
is the deeper target.  \emph{Provenance: article; claim-inventory paragraph 3.}
\end{definition}

\begin{claim}[C4: Multiplayer context shift]
Let $\mathrm{Vis}(c)$ be the collaborators who can observe context $c$.
Moving an agent from private context $c_p$ to shared channel $c_s$ with
$|\mathrm{Vis}(c_s)|>|\mathrm{Vis}(c_p)|$ changes the coordination pattern:
prompts, results, and relevant history become common process objects rather
than private assistant state.  \emph{Provenance: article; claim-inventory
paragraph 4.}
\end{claim}

\begin{definition}[C5: Independent service principal]
An agent identity $\iota_a$ is an independent service principal iff
$\iota_a\neq\iota_u$ for every human user $u$, and authorization is evaluated
as $\mathrm{Allow}(\iota_a,r,o)$ rather than by impersonating a delegator.
Tag's identity model is presented as satisfying this permission pattern.
\emph{Provenance: article; claim-inventory paragraph 5.}
\end{definition}

\begin{definition}[C6: Platform-bound compartmentalization]
For platform $L$, an agent is $L$-bound when identity, memory, or scope is a
function of platform compartments: $(\iota_a,\mu_a,\sigma_a)=F_L(c_L)$.
Portability fails when no semantics-preserving map transports these objects
to another communication substrate.  The article treats Slack channels as
Tag's binding compartments.  \emph{Provenance: article; claim-inventory
paragraph 6.}
\end{definition}

\begin{conjecture}[C7: Opaque-memory risk]
Define inspectability $J(\mu)$ as the fraction of action-relevant memory whose
content, provenance, and influence can be queried.  Holding other controls
fixed, behavioral-risk exposure $R$ is conjectured to be nonincreasing in
$J$: $\partial R/\partial J\leq 0$.  Low $J$ can conceal stale context, bias,
or sycophantic reinforcement.  \emph{Provenance: article; claim-inventory
paragraph 7.}
\end{conjecture}

\begin{definition}[C8: Symbolic metagraph memory]
A symbolic metagraph memory is $M=(V,E,\lambda,\pi,\nu)$, with typed atoms
$V$, higher-order links $E$, labels/types $\lambda$, provenance $\pi$, and
revision history $\nu$.  It is inspectable when queries can recover each of
these components and trace selected action premises.  Hyperon AtomSpace is a
candidate substrate; superiority to opaque neural memory remains a design
hypothesis, not a theorem.  \emph{Provenance: article; claim-inventory
paragraph 8.}
\end{definition}

\begin{theorem}[C9: Identity-collapse destroys unique attribution]
Let contextual agents $a_1\neq a_2$ map through an integration
$f:A\rightarrow I$ to upstream identities.  If $f(a_1)=f(a_2)$ and the audit
event records only $f(a_i)$, then the originating contextual agent is not
uniquely recoverable from that event.
\end{theorem}
\begin{proof}[Proof sketch]
The recorded identity has at least two preimages, so its inverse is not a
function on that event.  Additional provenance could restore attribution,
but is absent by hypothesis.  \emph{Provenance: article; claim-inventory
paragraph 9.}
\end{proof}

\begin{definition}[C10: Boundary audit versus live verification]
A boundary auditor observes only inputs, outputs, and completed action logs.
A live verifier $v$ observes execution states $s_t$, checks invariants
$\Phi(s_t)$ and protocol transitions $P(s_t,s_{t+1})$, and may signal or block
violations before completion.  The article characterizes Tag as providing
the former but not the latter.  \emph{Provenance: article; claim-inventory
paragraph 10.}
\end{definition}

\begin{conjecture}[C11: Tokenmaxxing incentive]
Let provider revenue be $r(q)=pq$ for metered token use $q$, while user utility
is $u(q)-pq$ with diminishing returns in $u$.  A product policy that defaults
to no budget constraint and benefits from $r'(q)>0$ does not by itself
implement the user's resource-minimizing optimum; explicit budgets or aligned
delegated cost policies are required.  \emph{Provenance: article;
claim-inventory paragraph 11.}
\end{conjecture}

\begin{claim}[C12: Open compositional sufficiency]
Let modules provide persistent agent loops, model bridges, chat transports,
and integration glue.  If their interfaces compose to realize identity,
shared dispatch, tools, and persistence, then an open ecosystem can reproduce
much of Tag's core functional envelope.  This is an architectural existence
claim, not an assertion of equal polish or operational quality.
\emph{Provenance: article; claim-inventory paragraph 12.}
\end{claim}

\begin{definition}[C13: Trustworthiness bottleneck triad]
The hard trust kernel is
$H=\langle M_{\mathrm{durable}},R_{\mathrm{isolated}},I_{\mathrm{audit}}\rangle$:
memory that remains correct and inspectable over time, runtime capabilities
confined by enforceable policy, and identity events with end-to-end
attribution.  Interface-level feature parity does not imply realization of
$H$.  \emph{Provenance: article; claim-inventory paragraph 13.}
\end{definition}

\begin{definition}[C14: OmegaClaw principal]
An OmegaClaw principal is
$\omega=\langle n,r,T,M,\rho,P\rangle$, comprising stable name, role, tools,
memory, cognitive remit, and permissions.  Persistence is attached to
$\omega$, while channels are replaceable interaction processes.
\emph{Provenance: article; claim-inventory paragraph 14.}
\end{definition}

\begin{conjecture}[C15: Reflective symbolic cognition path]
If an OmegaClaw's memory exports typed, provenance-bearing AtomSpace atoms and
its own role, beliefs, and actions are represented in the same metagraph,
then PLN, NARS-like revision, and self-model queries can operate on inspectable
cognitive objects.  Whether this yields robust cognition is an open empirical
question.  \emph{Provenance: article; claim-inventory paragraph 15.}
\end{conjecture}

\begin{definition}[C16: OmegaHive system]
An OmegaHive is a governed multi-agent process
$\mathcal{H}=\langle A_c,A_s,Q,G,L,P,V\rangle$, with cognitive agents $A_c$,
service agents $A_s$, task board $Q$, promotion rules $G$, permission lattice
$L$, inter-agent protocols $P$, and live verifier set $V$.  Emergent research
behavior is constrained by these explicit institutional objects.
\emph{Provenance: article; claim-inventory paragraph 16.}
\end{definition}

\begin{theorem}[C17: Portability under explicit agent state]
Model an agent society as explicit objects
$X=\langle A,M,P,O,C\rangle$.  Suppose identity, memory, permissions, and
observations are represented independently of communication context $C$, and
there exists a semantics-preserving transport adapter $g:C\to C'$.  Then the
society's agent-level state can be transported from $C$ to $C'$ without
redefining those four object classes.
\end{theorem}
\begin{proof}[Proof sketch]
Apply $g$ only to send/receive operations.  By independence, $A,M,P,O$ are
invariant; hence their relations and histories are preserved modulo transport
event renaming.  This formalizes the Hyperseed contrast between a
platform-bound role-pattern and a portable society of explicit cognitive and
process objects.  \emph{Provenance: article; claim-inventory paragraph 17.}
\end{proof}

\section{Open questions}
\begin{enumerate}
\item What is the minimal formal interface between Hyperseed concepts and AtomSpace/PLN atoms for this article series?
\item How should source-paraphrase claims be truth-valued separately from formalized design principles?
\item Which OmegaHive constructs should be modeled as agents, which as institutions, and which as protocol constraints?
\item How should resource-economics alignment be represented: as a value relation, a policy constraint, or a utility/attention budget?
\end{enumerate}

\end{document}
