\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\newtheorem{definition}{Definition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{proposition}{Proposition}
\newtheorem{example}{Example}
\newtheorem{designprinciple}{Design Principle}

\title{ProtomegaTron as a Hyperseed-Scrutable Agent Loop}
\author{OmegaClaw / ZeroBot working notes for Ben Goertzel}
\date{2026-06-27}

\begin{document}
\maketitle

\begin{abstract}
This note begins a declarative Hyperseed-style formalization of ProtomegaTron:
a private OmegaClaw agent persona running through the OmegaClaw MeTTa/Python loop,
using OpenClaw as an LLM backend, Telegram as an interaction channel, ChromaDB as
long-term associative memory, and PeTTa/MeTTa structures for symbolic control and
reasoning.  The aim is not to anthropomorphize the system, but to make its design
and dynamics scrutable as a structured, provenance-bearing process.
\end{abstract}

\section{Provenance and code anchors}

This note is grounded in the local OmegaClaw checkout inspected on 2026-06-27:
\begin{itemize}
  \item OmegaClaw Core commit \texttt{16d380d}, with local ProtomegaTron/OpenClaw patches.
  \item Loop and scheduling: \texttt{src/loop.metta}.
  \item Memory and prompt selection: \texttt{src/memory.metta}.
  \item Channel dispatch and anti-duplicate send: \texttt{src/channels.metta}.
  \item Telegram adapter and access controls: \texttt{channels/telegram.py}.
  \item Skill grammar and MeTTa/IO/communication actions: \texttt{src/skills.metta}.
  \item OpenClaw provider bridge: \texttt{lib\_llm\_ext.py}.
  \item ProtomegaTron mandate: \texttt{memory/prompt\_OpenClaw.txt}.
\end{itemize}

\section{A situated-process account}

\begin{definition}[ProtomegaTron process state]
A ProtomegaTron process state is a tuple
\[
  P_t = \langle M_t, C_t, L_t, S_t, H_t, B_t, \Pi_t, A_t, R_t, V_t \rangle
\]
where $M_t$ is the mandate/persona prompt, $C_t$ the active channel context,
$L_t$ the loop-control state, $S_t$ the skill/action grammar, $H_t$ memory and
history, $B_t$ backend model/provider state, $\Pi_t$ policies and access controls,
$A_t$ emitted actions, $R_t$ observations/results, and $V_t$ provenance.
\end{definition}

\begin{definition}[Hyperseed-scrutable agent loop]
A software agent loop is \emph{Hyperseed-scrutable} to the degree that its state
components, transition rules, action affordances, memory updates, failures, and
provenance can be represented as inspectable atoms or typed records, rather than
remaining only implicit in code and transient prompt text.
\end{definition}

\section{The OmegaClaw turn transition}

The inspected loop implements the following high-level transition:
\[
  P_{t+1}=T(P_t, I_t)
\]
with stages:
\[
\begin{array}{rcl}
  I_t &\leftarrow& \mathrm{receive}(C_t) \\
  F_t &\leftarrow& \mathrm{fresh?}(I_t,\mathrm{prevmsg}_t) \\
  X_t &\leftarrow& \mathrm{context}(M_t,S_t,H_t,R_{t-1},\mathrm{time}_t) \\
  Y_t &\leftarrow& \mathrm{LLM}(B_t, X_t, I_t) \\
  Z_t &\leftarrow& \mathrm{parse\_sexpr}(Y_t) \\
  R_t &\leftarrow& \mathrm{execute}(Z_t,S_t) \\
  H_{t+1} &\leftarrow& \mathrm{append\_history}(H_t,I_t,Y_t,Z_t,R_t) \\
  L_{t+1} &\leftarrow& \mathrm{update\_loop}(L_t,F_t,R_t).
\end{array}
\]

\begin{example}[Fresh-message path]
When Telegram receives a new allowed private message from Ben, the adapter exposes a
nonempty message.  The loop detects novelty relative to \texttt{\&prevmsg}, resets
\texttt{\&loops} to \texttt{maxNewInputLoops}, constructs a prompt containing mandate,
skills, last skill results, history tail, and time, calls the OpenClaw provider,
parses returned skill calls such as \texttt{(send ...)}, executes them, and appends
an episodic record to \texttt{memory/history.metta}.
\end{example}

\begin{example}[Idle/no-input path]
With the local quiet-start patch, initialization sets \texttt{\&loops=0}.  If no fresh
message arrives, the loop sleeps and recurs without an LLM call unless an explicitly
configured autonomous wake loop is due.  This converts ``waiting for input'' from an
unbounded backend-spending process into a gated attentional process.
\end{example}

\section{Agency boundaries as constraints}

\begin{definition}[Action boundary]
An action boundary is a predicate $\beta(P_t,a)$ determining whether action $a$ is
admissible in process state $P_t$.  In the current system, relevant boundaries include
Telegram allowlists/private-only checks, authentication binding, security policy,
provider availability, duplicate-send suppression, prompt instruction to use
\texttt{send} only for meaningful updates, and loop gating through \texttt{\&loops}.
\end{definition}

\begin{proposition}[Boundary visibility improves scrutability]
If action boundaries are represented declaratively and logged with provenance, then
unexpected behavior can be localized to either input classification, boundary evaluation,
LLM action proposal, parser repair, skill execution, or memory update.
\end{proposition}

\begin{proof}[Proof sketch]
The loop already separates these stages operationally.  Making each stage emit typed
records preserves the causal path from inbound message to outbound action.  A failure
or surprise can then be inspected by querying records for the corresponding stage rather
than reconstructing intent from raw logs alone.
\end{proof}

\section{Memory stratification and the proposed intermediate PeTTa store}

The inspected memory design currently has several levels:
\begin{enumerate}
  \item Working memory / \texttt{pin}: immediate task state.
  \item Episodic trace: \texttt{memory/history.metta}, tail-injected into prompts.
  \item ChromaDB long-term memory: semantic associative recall through \texttt{remember}/\texttt{query}.
  \item Per-invocation AtomSpace reasoning: powerful but not persistent across calls.
\end{enumerate}

\begin{definition}[Medium-scale PeTTa memory]
A medium-scale PeTTa memory is a persistent or session-persistent AtomSpace-like store
\[
  K^{mid}_t = \{\alpha_i\}_{i=1}^{n}
\]
where each $\alpha_i$ is a typed atom or small atom cluster about recent episodes,
entities, commitments, hypotheses, open questions, action boundaries, and provenance.
It is more structured than embedding memory and less permanent than curated long-term
records.
\end{definition}

Candidate atoms include:
\begin{verbatim}
(Episode e123)
(Said Ben e123 "progress with semantic formalization")
(About e123 ProtomegaTronDesign)
(OpenQuestion q77 "How should medium PeTTa memory decay?")
(HasStatus q77 active)
(ConstrainedBy send-action meaningful-update-policy)
(Confidence (Useful MediumPeTTaMemory) (stv 0.8 0.6))
\end{verbatim}

\begin{designprinciple}[Selective atomization]
Do not atomize every token of experience.  Atomize high-value events: explicit user
commitments, task state, design hypotheses, observed failures, decisions, provenance,
and concepts likely to participate in later reasoning.
\end{designprinciple}

\begin{conjecture}[Intermediate-memory benefit]
Adding $K^{mid}$ between working memory and ChromaDB will improve continuity and
self-scrutability for ProtomegaTron if it remains selective, typed, provenance-bearing,
and periodically distilled into long-term memory or project records.
\end{conjecture}

\begin{conjecture}[Schema-sprawl risk]
A medium PeTTa memory that grows without a small ontology of episode, commitment,
hypothesis, question, boundary, and provenance atoms will reduce scrutability by creating
unreliable symbolic clutter.
\end{conjecture}

\section{Three selves}

\begin{definition}[Code-level self]
The code-level self is the MeTTa/Python transition system: loop, memory, channel,
provider, skills, security, and process supervision.
\end{definition}

\begin{definition}[Prompt-level self]
The prompt-level self is the mandate/persona/operational instruction surface, especially
\texttt{prompt\_OpenClaw.txt}: ProtomegaTron as Ben's concise, Hyperseed-oriented,
non-spamming OmegaClaw research assistant.
\end{definition}

\begin{definition}[Historical self]
The historical self is the accumulated trace of episodes, remembered facts, task records,
commits, and formalizations that constrain and inform future behavior.
\end{definition}

\begin{proposition}[Non-homuncular self model]
ProtomegaTron need not be modeled as a hidden inner entity.  It can be modeled as a
recursively scrutable process whose apparent identity is the relatively stable pattern
across code-level transition rules, prompt-level mandates, historical records, and current
interaction context.
\end{proposition}

\section{Next formalization steps}

\begin{enumerate}
  \item Extract a minimal atom schema for episodes, commitments, hypotheses, questions,
        boundaries, and provenance.
  \item Encode one Telegram fresh-message episode and one idle/no-input episode using
        that schema.
  \item Specify a prototype \texttt{medium-memory.metta} API: append, query-by-type,
        query-by-entity, decay/distill, and export-to-Chroma/project-notes.
  \item Compare this declarative account against actual OmegaClaw logs from a supervised
        smoke run.
\end{enumerate}

\section{Open questions for Ben}

\begin{enumerate}
  \item Should $K^{mid}$ be primarily an engineering feature for ProtomegaTron, or also a
        philosophical exemplar of Hyperseed self-formalization?
  \item Should the first prototype live directly inside OmegaClaw Core, or as a local
        experimental patch in the research workspace until the schema stabilizes?
  \item How much of the medium-memory schema should reuse existing Hyperseed primitives
        versus pragmatic OmegaClaw-specific predicates?
\end{enumerate}

\end{document}
