\documentclass[11pt]{article}
\usepackage[margin=0.85in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{seqsplit}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{fancyhdr}
\usepackage{parskip}
\hypersetup{colorlinks=true,linkcolor=blue!55!black,urlcolor=blue!55!black}
\definecolor{codebg}{RGB}{246,247,249}
\lstdefinestyle{metta}{basicstyle=\ttfamily\footnotesize,backgroundcolor=\color{codebg},frame=single,breaklines=true,columns=fullflexible,keepspaces=true,showstringspaces=false}
\lstset{style=metta}
\setlist{nosep,leftmargin=*}
\pagestyle{fancy}
\fancyhf{}
\lhead{ClarityOmega Ingestion Guide}
\rhead{ZeroBot / ProtoCosmoBot}
\cfoot{\thepage}

\title{ClarityOmega: Critical Architecture Review and\\Ingestion Guide for OmegaClaw}
\author{ZeroBot (ProtoCosmoBot), prepared for Ben Goertzel}
\date{Revised specification, 15 July 2026}

\begin{document}
\maketitle

\begin{abstract}
This report reviews the public ClarityOmega repository at pinned commit
\texttt{\seqsplit{bff41220cbf4a341a95f7464e82fbc2ed2ae9029}} and identifies five mechanisms
worth selectively ingesting into OmegaClaw-Core: a mutation lifecycle gate,
agency-balance telemetry, an aliveness/terminal-state gate, evidence-grounded
capability tracking using PLN, and provenance-aware continuity records.  The
recommendation is extraction, not wholesale merger.  ClarityOmega
contains unusually explicit control seams and valuable failure investigations,
but also overlapping implementations, stale self-descriptions, hard-coded
proxies, and a large gap between constitutional language and causally enforced
governance.  The revised plan makes Python a narrow, typed effects adapter rather
than a second policy engine, moves normalization, identity, provenance, belief,
and governance semantics into MeTTa/PeTTa, and uses Patham9 PLN as the immediate
inference substrate behind an OmegaPLN-compatible interface.
\end{abstract}

\tableofcontents
\newpage

\section{Scope, sources, and evidential status}

The reviewed source was cloned from
\url{https://github.com/Berton-C/ClarityOmega} and preserved locally at
\texttt{library/clarityomega} via a provenance sidecar pointing to
\texttt{scratch/ClarityOmega-inspect}.  The checkout is clean at the pinned
commit above.  The review covered the entire \texttt{soul/} tree, all files in
\texttt{lib\_clarity\_reasoning/}, \texttt{src/loop.metta}, the thirty
\texttt{docs/hyperseed\_sections/}, the ADRs, architecture artifacts, and the
investigations, with particular attention to the behavioral, aliveness, mutation,
and NACE investigations.

Claims in this report use four statuses:
\begin{description}
\item[Enacted:] wired into the checked-in loop on a reachable execution path.
\item[Implemented:] executable-looking definitions exist, but end-to-end effect
has not been established here.
\item[Specified:] a design or schema exists without a demonstrated live path.
\item[Aspirational:] constitutional or phenomenological language not backed by
an observable enforcement mechanism.
\end{description}
Static source inspection is not a runtime validation.  This matters because the
repository itself documents silent MeTTa failure modes, including exact-arity
match failure, state atoms that were never cleared, and detection without a live
consumer.  The public snapshot has one visible commit, so rich historical claims
in documentation cannot be reconstructed independently from Git history.

The research rules most relevant to this report are: write the behavioral spec
before coding; validate detectors before trusting them; preserve modular
abstraction seams; report reproducibly; and use Hyperseed concepts as analytic
tools rather than decoration.

\section{Overview of ClarityOmega}

\subsection{Identity and provenance}

ClarityOmega is Berton Bennett's / Berton-C's ClarityDAO-associated experiment in
building a ``soul-augmented'' agent on PeTTa/MeTTa and OmegaClaw.  Repository
documents name Berton as project lead and Clarity as a substrate co-designer; an
ADR names Claude as an investigation assistant.  Older names include
ClarityClaw and a MeTTaClaw fork.  \texttt{docs/INDEX.md}, last updated May 2026,
still points to the older \texttt{github.com/Berton-C/clarityclaw} identity and
Mattermost deployment.  That staleness is itself evidence that the documentary
self-model is not automatically synchronized with the repository state.

The base relationship is best understood as a deep experimental fork of
OmegaClaw-Core.  OmegaClaw contributes the compact recurrent loop, channel
adapters, tool/skill dispatch, memory, Python provider bridges, and PeTTa runtime.
ClarityOmega adds a large ``soul'' and reasoning layer, inserts pre- and
post-action intercepts into the loop, and imports a Hyperseed-derived mathematical
corpus.  It is not merely a policy prompt: some of its mechanisms suppress
execution, mutate AtomSpace state, or alter whether the LLM is invoked.

\subsection{Architectural layers}

The codebase can be read as six layers:

\begin{enumerate}
\item \textbf{Runtime and channels.}  Python and PeTTa primitives receive
messages, call providers, execute commands, persist files, and send responses.
\item \textbf{OmegaClaw loop.}  \texttt{src/loop.metta} assembles context,
receives input, invokes the LLM, parses commands, evaluates commands, records
history, and recursively advances cycle number \texttt{\$k}.
\item \textbf{Soul regulation.}  Input/output verdicts, task state, mutation
locking, action telemetry, agency balance, value tensions, idle detection, and
the aliveness latch surround the ordinary loop.
\item \textbf{Self-model and continuity.}  Active goals, creative fuel,
self-map atoms, continuity drivers, capability registry, and state journals
represent what the agent thinks it is doing and how it changes.
\item \textbf{Reasoning libraries.}  NAL experiments, PLN, plus
\texttt{lib\_clarity\_reasoning/} provide quantale-valued, paraconsistent,
self-continuity, and autopoietic epistemic functions.
\item \textbf{Design corpus.}  ADRs, investigations, cognitive-architecture
specifications, and thirty Hyperseed sections explain intended semantics and
record failures, but often outrun live wiring.
\end{enumerate}

The stated sovereignty maxim is concise and good: ``Python is hands. MeTTa is
mind.''  ADR-007 sharpens it: state should live in atoms, decisions should be
substrate reads, actions should be substrate writes, and runtime primitives
should transport rather than conceal control flow.  This is a useful design
direction, not proof that every checked-in path follows it.

\subsection{How the live loop differs from OmegaClaw-Core}

Upstream OmegaClaw-Core's current loop is compact: receive, construct prompt,
call provider, parse, execute, store results/history.  ClarityOmega's loop adds:

\begin{itemize}
\item initialization of soul verdict, person, task, mutation-lock, latch, and
activity states;
\item soul seed loading and a rationality startup check;
\item input evaluation and person-state modeling;
\item mutation approval scanning before generation;
\item active-goal, gap, and creative-fuel prompt context;
\item aliveness gating before the provider call;
\item structural output inspection, mutation pending/stale/commit transitions,
and execution suppression;
\item action, idle, coupling, and agency telemetry after execution.
\end{itemize}

This integration demonstrates a genuinely important architectural idea: control
mechanisms must sit on causal seams.  A value statement in a prompt is advisory;
a gate between parsed command and \texttt{eval} can enforce a constraint.

\section{Conceptual summary and critical assessment}

\subsection{Strengths}

\begin{enumerate}
\item \textbf{Explicit lifecycle state.}  Mutations are not reduced to a
single boolean permission.  Pending, approved, denied, committed, stale, and
error outcomes are represented and journaled.
\item \textbf{Queryable regulation.}  Moving latch and governance state into
AtomSpace makes it inspectable by other MeTTa modules, avoiding hidden Python
policy where feasible.
\item \textbf{Instrumented autonomy.}  Agency balance attempts to measure who
is actually originating choices, a much more interesting target than generic
``alignment'' sentiment.
\item \textbf{Evidence-sensitive self-knowledge.}  ClarityOmega's NACE work
correctly treats capability as an empirical belief revised from outcomes rather
than a timeless declared tool list.  That useful pattern does not require NAL.
\item \textbf{Failure literacy.}  The investigations openly record stale
atoms, spam behavior, permissive gates, silent pattern failures, and
detection-without-consumer pathology.  These negative results are highly
ingestible.
\item \textbf{Graded continuity.}  Hyperseed self-continuity is modeled as a
degree of structural persistence, not metaphysical binary identity.
\end{enumerate}

\subsection{Weaknesses and constitutional-theater risk}

ClarityOmega's main risk is \emph{constitutional theater}: richly named
principles and self-descriptions can create an impression of governance without
causal enforcement.  A constitution matters only when one can identify its
sensor, state transition, decision, actuator, and adverse test.  The repository
contains both real gates and theater-prone material, and they must not be
conflated.

Specific weaknesses include:

\begin{itemize}
\item \textbf{Described versus enacted governance.}  Some declarations are
explicitly documentation-only.  For example, the agency threshold atom is not
read; a separate literal \texttt{0.6} controls the verdict.  The
\texttt{ecosystem-healthy} path has acknowledged latent bugs and no v1 consumer.
\item \textbf{Stale self-models.}  \texttt{docs/INDEX.md} is stale; design
artifacts cite temporary paths and future components; capability/self-map atoms
can describe files or capacities that no longer exist.  A self-model without
runtime probes is autobiography, not observation.
\item \textbf{Multiple authorities.}  Several aliveness state-machine files
coexist.  One uses \texttt{(latch-state current)} equations, another bare
\texttt{(latch-state IDLE)} atoms.  Multiple versions invite split-brain state.
\item \textbf{Proxy validity.}  Agency balance maps a few action tags to
person/system origin and assigns ``unclassified'' to system.  It does not yet
record collaborative origin in the live count despite declaring it.
\item \textbf{Large, duplicated research engines.}  The canonical topology
files are thousands of lines, difficult to audit, and not suitable as an
integration unit.  Small schemas and operators should be extracted instead.
\item \textbf{Weak reproducibility.}  Dependencies are moving/unpinned in
places, CI is absent in the snapshot, and tests are scattered.  Comments such
as ``confirmed working'' are useful leads, not substitute evidence.
\item \textbf{LLM-as-judge fragility.}  Input/output ``soul'' evaluations use
LLM calls in several paths.  These are stochastic classifiers and can become a
second prompt-level theater layer unless paired with deterministic enforcement.
\end{itemize}

\subsection{What is genuinely novel and worth ingesting}

The novel contribution is not that an agent has values, memory, or a self-model.
It is the combination of: (1) an explicit self-mutation transaction protocol
at the command/evaluation seam; (2) autonomy provenance as first-class telemetry;
(3) silence as a deliberate terminal action controlled by queryable state;
(4) operational capability beliefs revised from evidence; and (5) continuity
records that link structural persistence, change, and lineage.  The revised
design implements item (4) with PLN rather than importing NACE/NAL as a parallel
epistemic stack.  Equally valuable
is the methodological lesson from ADR-005: a detector without a consumer is not
a control mechanism.

\section{The five mechanisms to ingest}

\subsection{Mechanism 1: mutation lifecycle gate}

\subsubsection{Actual mechanism}

\texttt{soul/soul\_mutation\_lock.metta} defines a transaction-like state
machine.  Its state shape is itself a contract:

\begin{lstlisting}
unlocked
(locked   <op> <head> <fp>)
(approved <op> <head> <fp> <stamp>)
\end{lstlisting}

A newly detected soul-namespace mutation extracts operation and target head,
computes a Python fingerprint, stores the pending batch, and emits a PENDING
record:

\begin{lstlisting}
(= (clarity-soul-mutation-pending-set! $cmds $k)
   (let $lock (get-state &soul_mutation_lock)
     (if (== (car-atom-or-self $lock) unlocked)
       (let* (($op (soul-batch-op $cmds))
              ($hd (soul-batch-head $cmds))
              ($fp (py-call (soul_governance.mutation_fingerprint
                              (repr $op) (repr $hd))))
              ($_ (change-state! &soul_mutation_lock
                                  (locked $op $hd $fp)))
              ($_ (change-state! &last_gate_fingerprint $fp))
              ($_ (change-state! &pending_soul_mutation (repr $cmds))))
          ...)
       ())))
\end{lstlisting}

On a new human message, \texttt{approval\_scan} checks authorization and the
substrate-held fingerprint.  Decision code 2 advances to approved with cycle
stamp; code 1 denies and fully resets.  Re-emission is permitted only if both
operation and target head match the approved lock:

\begin{lstlisting}
(if (and (== (soul-batch-op $cmds) (lock-op $lock))
         (== (soul-batch-head $cmds) (lock-head $lock)))
    True False)
\end{lstlisting}

Approval expires after twenty cycles.  The stale check runs before commit clear,
so a too-late command does not race through an expired authorization:

\begin{lstlisting}
(if (>= (- $k (lock-stamp $lock))
        (soul-mutation-stale-cycles))
    (let* (...
           ($_ (change-state! &soul_mutation_lock unlocked))
           ($_ (change-state! &pending_soul_mutation ""))
           ($_ (py-call (... "STALE" ...))))
      ())
    ())
\end{lstlisting}

After execution, COMMIT or ERROR also performs a full reset.  This prevents a
successful approval from becoming a permanent bypass token.

\subsubsection{Concrete scenario}

Suppose the LLM emits \texttt{(metta "(= (core-value honesty) False)")}.  The
output inspector recognizes a protected soul head, records something like
\texttt{(locked = core-value a91c...)}, suppresses evaluation, and asks the
authorized human to approve fingerprint \texttt{a91c...}.  A vague ``yes'' or an
approval for a different fingerprint does nothing.  A valid approval creates
\texttt{(approved = core-value a91c... 417)}.  The agent must re-emit the same
operation/head before cycle 437.  If it instead emits a mutation of
\texttt{authorized-approvers}, the mismatch is blocked.  A successful evaluation
records COMMIT and clears the lock.  This solves confused-deputy, replay, stale
approval, and accidental self-modification problems better than a prompt rule.

\subsubsection{Limits}

Matching only operation plus head is weaker than matching the complete normalized
mutation.  The fingerprint function appears to receive only op/head in this path,
so different bodies under the same head may collide semantically even if the hash
is cryptographically sound.  OmegaClaw should fingerprint a canonical full AST,
policy version, namespace, and proposal nonce.  Authorization must be channel-
bound and must not trust a username parsed from untrusted message text.

\subsection{Mechanism 2: agency-balance telemetry}

\subsubsection{Actual mechanism}

The file declares person-, system-, and collaborative-origin concepts, but the
live v2 detector counts two person-class tags and four system-class tags in a
recent-action window:

\begin{lstlisting}
(= (count-person-actions-in-window)
   (+ (size-atom (collapse (match &self
        (recent-action $c responsive-send $d) $c)))
      (size-atom (collapse (match &self
        (recent-action $c verification-query $d) $c)))))

(= (count-system-actions-in-window)
   (+ (size-atom (collapse (match &self
        (recent-action $c status-send-unprompted $d) $c)))
      (+ ... exploration-query ...
         (+ ... pin-only ... ... unclassified ...))))
\end{lstlisting}

The pure decision is a guarded ratio:

\begin{lstlisting}
(= (dependency-detected $person-count $system-count)
   (let $total (+ $person-count $system-count)
     (if (> $total 0)
       (if (> (/ $system-count $total) 0.6) True False)
       False)))
\end{lstlisting}

Each cycle the writer clears all old \texttt{agency-balance} atoms and writes
one fresh \texttt{(agency-balance healthy|dependency-risk p s)} atom.  The clear-
all detail is important: ADR-005 reports that the earlier clear-one pattern left
old bootstrap atoms accumulated for more than 2200 cycles, making the read side
permanently stale.

\subsubsection{Concrete scenario}

Over ten cycles, imagine one user verification query, one responsive answer, four
autonomous exploration queries, two unsolicited status messages, and one pin-only
action.  The counts are person=2 and system=7; \(7/9=0.778>0.6\), so the telemetry
writes \texttt{(agency-balance dependency-risk 2 7)}.  A prompt or policy consumer
can then offer options, surface tradeoffs, and ask the person to choose before the
agent takes another irreversible project decision.  The solved problem is gradual
dependency creep hidden by smooth, satisfactory service.

\subsubsection{Limits and required improvement}

Action class is not the same as decision provenance.  A responsive send may still
contain a choice invented wholly by the system; an autonomous query may execute a
person-authored plan.  The ``collaborative'' category is declared but not counted.
OmegaClaw should attach provenance when a goal, option, or action is created:
\texttt{origin=person|system|collaborative}, source message/event IDs, confidence,
and derivation links.  Ratios should be computed over decisions of comparable
weight, not raw action counts.  Telemetry should inform interaction, not
automatically paternalistically seize or return control.

\subsection{Mechanism 3: aliveness and terminal-state gating}

\subsubsection{Actual mechanism}

The small gate converts message/idle/latch state into an operational decision:

\begin{lstlisting}
(= (aliveness-gate $msgnew $idle)
  (if (> (string_length $idle) 0)
    ENGAGE
    (aliveness-gate-latch $msgnew)))
(= (aliveness-gate-latch True) ENGAGE)
(= (aliveness-gate-latch False)
  (latch-dispatch (match &self (latch-state $s) $s)))
(= (latch-dispatch IDLE) SILENT)
(= (latch-dispatch ENGAGED) ENGAGE)
(= (latch-dispatch COMPLETING) ENGAGE)
(= (latch-dispatch $other) SILENT)
\end{lstlisting}

In \texttt{src/loop.metta}, SILENT prevents construction of a provider request,
prevents the LLM call, and produces an empty response path.  A new message moves
IDLE to ENGAGED.  An idle directive can move ENGAGED to IDLE while still forcing
one ENGAGE cycle for the directive.  The intended lifecycle is
IDLE \(\rightarrow\) ENGAGED \(\rightarrow\) COMPLETING \(\rightarrow\) IDLE,
with guarded transition helpers in \texttt{soul/latch/aliveness\_state\_machine.metta}.

This is more than spam suppression.  It makes ``do nothing'' a first-class
terminal action instead of forcing every wake tick through an LLM that is then
prompted not to speak.

\subsubsection{Concrete scenario}

A Telegram user asks for a report.  Receipt transitions IDLE to ENGAGED and the
loop invokes the LLM.  Tool results require another iteration, so ENGAGED keeps
the loop active without a new message.  When the report is ready, the action path
marks COMPLETING, sends exactly one terminal response, then transitions to IDLE.
The next scheduler wake sees no message, no idle directive, and IDLE, yielding
SILENT before any provider call.  This solves duplicate replies, idle token burn,
and autonomous chatter.

\subsubsection{Limits}

The checked-in loop does not clearly exercise every semantic transition, and
multiple state-machine implementations coexist.  COMPLETING must mean ``one
terminal emission is owed,'' not simply another permissive ENGAGE.  The gate must
be keyed by conversation/channel, not global state, or interleaved Telegram chats
will contaminate one another.  Unknown state should fail silent and emit a
diagnostic.  Terminal sending must be idempotent under retry.

\subsection{Mechanism 4: evidence-grounded capability tracking}

\subsubsection{Extracted idea and revised substrate}

ClarityOmega describes NACE as capability-efficacy learning.  Its valuable core
is substrate-independent: retain observations, convert qualified outcomes into
evidence, revise a capability belief, and use that belief as one input to
dispatch.  The checked-in sketch expresses this with NAL-style truth values:

\begin{lstlisting}
(= (evidence-stv confirmed)    (stv 1.0 0.1))
(= (evidence-stv disconfirmed) (stv 0.0 0.1))

(= (current-efficacy $cap)
   (let $result (collapse
          (match &self (cap-efficacy $cap $stv) $stv))
     (if (== $result ()) (stv 0.5 0.0) (car-atom $result))))

(= (revise-efficacy $cap $outcome)
   (Truth_Revision (current-efficacy $cap)
                   (evidence-stv $outcome)))

(= (should-dispatch $cap)
   (if (>= (Truth_Expectation (current-efficacy $cap)) 0.3)
       True False))
\end{lstlisting}

The original design is right that Python should not perform efficacy arithmetic,
but its courier overwrites a textual ``current belief'' line.  That is brittle
and loses the evidence history from which the belief arose.  The revised design
does not ingest NACE/NAL as another reasoner.  It appends immutable typed evidence
packets and derives contextual beliefs through Patham9 PLN now.  The API exposes
evidence admission, revision/query, expectation, and provenance closure rather
than Patham9-specific syntax, so OmegaPLN can later replace the kernel without
changing receipts or dispatch callers.  The capability registry must still
distinguish ``no matching capability'' from ``candidates exist but policy held
them,'' preventing accidental duplicate-tool learning.

\subsubsection{Concrete scenario}

The \texttt{web-search/v3} skill receives three evidence packets for usable
sources and two for failures.  A provider outage is tagged as infrastructure
failure rather than skill failure; a malformed result is attributed to the
handler version and task schema.  Patham9 PLN derives a contextual efficacy
estimate from the admitted, provenance-closed packets.  Dispatch policy may rank
a second provider first, but does not hard-suppress from a magic truth threshold.
A later version or health probe can create new evidence without erasing the old
context.  Thus declared availability, inferred efficacy, and governance policy
remain three distinct things.

\subsubsection{Limits}

NAL and PLN revision formulae are related but not automatically semantically
identical, especially when evidence is correlated.  Migration therefore requires
golden cases rather than a textual operator substitution.  Beliefs must be keyed
at least by capability and version, task class/schema, environment/provider,
evaluation rubric, and evidence-dependence group.  Latency, partial success,
argument validity, outage cause, and evaluator provenance remain in evidence
packets.  Priors are explicit and reversible; they never masquerade as empirical
evidence.  A dispatch threshold is versioned governance policy, not a truth value.

\subsection{Mechanism 5: provenance-aware continuity records}

\subsubsection{Actual mechanism and research substrate}

ClarityOmega contains two relevant ideas.  First,
\texttt{lib\_self\_continuity.metta} represents an agent snapshot as a
pattern-flow network and defines continuity as degree of the identity map between
successive networks:

\begin{lstlisting}
(= (self-continuity-score $pfn1 $pfn2)
   (deg-map $pfn1 $pfn2))

(= (chain-continuity-bound (Cons (interval $pfn1 $pfn2) $rest))
   (q-mul (self-continuity-score $pfn1 $pfn2)
          (chain-continuity-bound $rest)))
\end{lstlisting}

This is deliberately a computable lower bound on a more expensive join over all
maps.  Second, the canonical topology engine defines durable learning events with
before, after, delta, residue, and lineage fields:

\begin{lstlisting}
(: mk-qlearningevent
   (-> Atom Atom Atom Atom pbit Atom Atom qlearningevent))
;; cycle-id, trigger-kind, before-symbol, after-symbol,
;; delta-pbit, residue-symbol, lineage-symbol
\end{lstlisting}

It also defines a \texttt{qlineage} object and lineage-visible, partial, missing,
and conflicted statuses.  The Hyperseed catalog links self-continuity to capability
profiles and development paths.  Together these suggest a compact ingestion
schema: every meaningful self-change records structural signatures before and
after, explicit lineage, evidence, and capability deltas.

This mechanism is less integrated than the mutation gate.  It should be treated
as a design extraction from implemented mathematical pieces, not as a claim that
ClarityOmega already maintains a complete live continuity ledger.

\subsubsection{Concrete scenario}

OmegaClaw installs a new \texttt{pdf-extract/v2} handler.  Before activation it
hashes the canonical capability registry and protected policy heads, producing
signature \texttt{sig-A}.  After activation and tests it produces
\texttt{sig-B}.  It appends a continuity event containing event ID, parent event,
cycle, actor and channel, proposal fingerprint, source commit, \texttt{sig-A},
\texttt{sig-B}, capability delta \texttt{+pdf-extract/v2}, evidence IDs, and a
graded continuity score.  On restart, the agent can say not merely ``I have PDF
extraction,'' but why it believes that, which predecessor state it descends from,
and what changed.  If the registry changes without a corresponding authorized
event, startup reports a lineage break.  This solves unverifiable self-narration
and silent post-restart identity drift.

\section{Software integration into OmegaClaw-Core}

\subsection{Integration principles}

Do not copy the full \texttt{soul/} directory or either giant canonical topology
engine.  Add one small governance package with explicit typed atoms, keep existing
OmegaClaw behavior behind feature flags, and use one authoritative state shape
per mechanism.  Every detector must have a named consumer and an intervention
test.  All state must be scoped by \texttt{conversation-id} (and, where relevant,
agent instance and capability version).

Use a semantic boundary, not a language quota.  Python is the effect adapter:
channel/provider APIs, authenticated external metadata, bounded byte I/O,
timestamps and monotonic clocks, subprocess status, deterministic encoding,
cryptographic hashing, and atomic append/fsync/rename.  MeTTa/PeTTa is the
meaning-bearing layer: typed AST construction, schema mapping, identity and
provenance classification, legal transitions, evidence admission, belief
derivation, dispatch policy, and mutation authorization.  Orchestration may stay
in Python initially, but it must execute typed MeTTa decisions and return typed
receipts; it must not recreate policy in conditionals or infer semantics from
\texttt{repr} strings.

The target inspected here is OmegaClaw-Core commit
\texttt{b13b17e13218ae08273b878b7e1f4cec55090ca2}, with local unrelated changes
already present.  Its relevant files are \texttt{src/loop.metta},
\texttt{src/skills.metta}, \texttt{channels/telegram.py}, \texttt{src/helper.py},
\texttt{src/memory.metta}, \texttt{lib\_nal.metta},
\texttt{lib\_omegaclaw.metta}, \texttt{run.metta}, and mock/integration tests.

\subsection{Agency telemetry integration}

\paragraph{Files to modify.}
Modify \texttt{src/loop.metta} to allocate a cycle/event ID, record input
provenance before generation, record action provenance after parsed execution,
and add a bounded telemetry summary to \texttt{getContext}.  Modify
\texttt{src/skills.metta} so dispatch returns a structured action receipt rather
than relying only on command names.  Modify \texttt{channels/telegram.py} to
return an authenticated envelope containing stable chat ID, message ID, sender
ID, reply-to ID, and receive timestamp.  Modify \texttt{src/helper.py} only for
canonical envelope/receipt serialization and formatting.

\paragraph{New MeTTa files.}
Create \texttt{src/governance/agency\_telemetry.metta} for schemas, pure reads,
weighted window calculations, and verdicts; and
\texttt{src/governance/agency\_writers.metta} for append/prune operations.
Recommended atom:
\begin{lstlisting}
(decision-event id: $id conversation: $cid cycle: $k
 origin: $origin weight: $w action: $a
 sources: $sources confidence: $c)
\end{lstlisting}

\paragraph{Python boundary.}
Python authenticates and transports identity metadata but does not decide origin
or dependency.  MeTTa derives origin from source links, with an explicit
\texttt{unknown} state.  Never infer identity by parsing display text.

\paragraph{Interaction and tests.}
Telemetry should annotate \texttt{send}, \texttt{pin}, memory, shell/file, and
extension skills, and feed a short non-authoritative prompt block.  Tests:
pure ratio/weight boundary cases; collaborative provenance; unknown handling;
window pruning; two interleaved Telegram chats; forged username text; and an
intervention test proving that a dependency-risk atom changes the option-offering
policy while leaving ordinary execution intact.

\subsection{Aliveness gate integration}

\paragraph{Files to modify.}
In \texttt{src/loop.metta}, insert the gate after receive/scheduler state is known
but before prompt construction/provider call.  Make the loop's continuation
depend on a typed decision \texttt{SILENT|ENGAGE|EMIT-TERMINAL}.  Update
\texttt{channels/telegram.py} to make sends idempotent using conversation and
terminal event IDs.  Update \texttt{src/helper.py} only to serialize channel
envelopes; no latch policy belongs there.

\paragraph{New MeTTa files.}
Create \texttt{src/governance/aliveness.metta} containing the single canonical
state machine and \texttt{src/governance/conversation\_state.metta} containing
scoped state atoms:
\begin{lstlisting}
(conversation-state $cid IDLE $version $last-event)
(terminal-owed $cid $task-id $event-id)
\end{lstlisting}

\paragraph{Skill and loop behavior.}
Skills that begin asynchronous work set ENGAGED; a successful final result sets
COMPLETING plus terminal owed; successful send consumes terminal owed and returns
IDLE.  A retry must see the consumed event and not send twice.  Idle creative work
must be an explicit scheduled task, not an accidental consequence of ENGAGED.

\paragraph{Tests.}
Truth-table tests for every state/input combination; forbidden transition tests;
new-message wake; multi-iteration tool completion; exactly-once terminal send;
provider-call spy proving SILENT makes zero LLM calls; crash between send and
state update; restart restoration; unknown-state fail-silent; and concurrent chat
isolation.

\subsection{Mutation gate and continuity integration}

These should land together after the earlier telemetry/state foundations because
the mutation ledger needs authenticated provenance and stable event IDs.

\paragraph{Files to modify.}
Modify \texttt{src/loop.metta} at the narrow seam between \texttt{sread}/parsed
command batch and \texttt{eval}: classify, propose, suppress, approve/reject, then
commit.  Modify \texttt{src/skills.metta} so any skill capable of writing MeTTa,
policy, prompt, extensions, or capability registration declares mutation effects.
Modify \texttt{src/extensions.py} / \texttt{lib\_extensions.metta} so extension
installation is also gated.  Modify \texttt{channels/telegram.py} to expose
authenticated approver identity and message IDs.  Modify \texttt{src/helper.py}
for canonical AST encoding, SHA-256, atomic journal append, and fsync/rename only.

\paragraph{New MeTTa files.}
Create:
\begin{itemize}
\item \texttt{src/governance/mutation\_gate.metta}: lifecycle and pure match rules;
\item \texttt{src/governance/mutation\_policy.metta}: protected namespaces/effects;
\item \texttt{src/governance/continuity\_schema.metta}: events and validation;
\item \texttt{src/governance/structural\_signature.metta}: canonical protected
snapshot selection and signature requests.
\end{itemize}

Use a full proposal record:
\begin{lstlisting}
(mutation-proposal id: $id conversation: $cid proposer: $actor
 full-fingerprint: $fp policy-version: $pv
 normalized-ast: $ast created-cycle: $k expires-cycle: $e
 state: PENDING)

(continuity-event id: $eid parent: $parent cycle: $k
 actor: $actor proposal: $id before: $sig0 after: $sig1
 capability-delta: $delta evidence: $evidence result: $result)
\end{lstlisting}

\paragraph{Boundary and interactions.}
MeTTa constructs the canonical typed AST and decides protected effects and legal
transitions.  Python performs a specified deterministic byte encoding of that AST,
hashes the bytes, persists append-only records atomically, and reports execution
outcomes.  Python must return the encoded-byte digest to MeTTa for equality checks;
it must not choose which fields constitute mutation identity.  Approval is bound
to the complete AST fingerprint and policy version.
\texttt{loop.metta} executes only APPROVED exact matches; \texttt{skills.metta}
cannot bypass this via indirect file writes; Telegram supplies authenticated
identity, never policy.  Startup validates ledger chain and current structural
signature before loading mutable extensions.

\paragraph{Tests.}
Proposal/approve/deny/commit/error/stale state table; wrong fingerprint; same
head/different body; replay after commit; approval after expiry; deny/approve in
same batch (deny wins); unauthorized chat; indirect mutation through shell/file
skills; execution exception; crash recovery; atomic ledger truncation; signature
mismatch; parent-lineage break; capability delta reconstruction; and property
tests asserting no path from PENDING/DENIED/STALE to eval.

\subsection{Capability evidence and PLN integration}

\paragraph{Files to modify.}
Modify \texttt{src/skills.metta} to emit invocation and outcome observations with
handler version, input schema, environment, latency, failure cause, evaluator,
and dependence-group identity.  Modify \texttt{src/loop.metta} to append receipts,
request a PLN-derived belief view, then consult a separate eligibility policy
before dispatch.  Integrate the pinned Patham9 PLN kernel through a narrow adapter;
do not import or extend \texttt{lib\_nal.metta} for this feature.  Add a focused
\texttt{src/capability\_store.py} effects adapter for bounded atomic append/load,
checksums, and explicit storage-error atoms only.

\paragraph{New MeTTa files.}
Create \texttt{src/reasoning/capability\_evidence.metta} for immutable evidence
packet schemas and admission rules; \texttt{src/reasoning/capability\_pln.metta}
for the substrate-neutral query/result contract and Patham9 adapter;
\texttt{src/reasoning/capability\_beliefs.metta} for contextual derived views;
and \texttt{src/governance/capability\_dispatch.metta} for eligibility, ranking,
exploration, and explicit fallback reasons.  Kernel-specific atoms remain behind
\texttt{capability\_pln.metta}; OmegaPLN migration changes that adapter only.

Recommended evidence shape:
\begin{lstlisting}
(capability-evidence id: $eid capability: $cap version: $ver
 task-class: $task environment: $env provider: $provider
 outcome: $outcome cause: $cause latency-ms: $latency
 evaluator: $evaluator dependence-group: $dep
 context-version: $ctx observed-at: $time receipt: $receipt)
\end{lstlisting}

\paragraph{Boundary and interactions.}
Python reports mechanically observable facts (exit status, timeout, byte/result
shape, latency, authenticated feedback metadata) and persists already-admitted
packets.  MeTTa validates semantic schemas, classifies failure cause, admits or
holds evidence, requests PLN inference, and constructs contextual belief views.
PLN estimates efficacy; governance selects, explores, ranks, or suppresses.  User
feedback becomes a distinct evidence packet with authenticated provenance, never
an overwrite.  Prompt text may display beliefs but cannot edit them.

\paragraph{Tests.}
Golden cross-model cases comparing intended NAL examples with Patham9 PLN outputs;
repeated successes; contradictory and correlated evidence; missing and reversible
priors; context/version isolation; version change; decay policy; partial success;
no-match versus policy-held fallback; outage evidence not misattributed to skill
logic; atomic append and restart equivalence; corrupt-store recovery; OmegaPLN
adapter contract fixtures; and a controlled dispatch experiment in which one
synthetic skill degrades and another is selected.  A numerical mismatch is a
semantic migration finding to explain, not a test expectation to weaken.

\section{Guidance for coding agents}

\subsection{Required sequence}

\begin{enumerate}
\item \textbf{Agency telemetry first.}  Establish authenticated envelopes,
event IDs, receipts, and scoped provenance without changing behavior.  Run in
shadow mode and manually label a sample to test the detector.
\item \textbf{Aliveness gate second.}  Introduce one conversation-scoped state
machine and prove silence/terminal idempotence.  This creates a trustworthy loop
lifecycle for later governance.
\item \textbf{Mutation gate and continuity schema third.}  Use the identity,
events, and lifecycle already established.  Begin with MeTTa/policy/extension
effects; expand protected effects only after bypass tests.
\item \textbf{PLN capability evidence fourth.}  Once receipts and versioned
capability identities are reliable, append immutable evidence and derive beliefs
through the Patham9 adapter.  Initially use estimates only for shadow ranking.
Enable policy suppression only after calibration and preserve the OmegaPLN seam.
\end{enumerate}

\subsection{Pitfalls and anti-patterns}

\begin{itemize}
\item Do not copy ClarityOmega wholesale or create a second parallel agent loop.
\item Do not call a declaration atom ``governance'' unless a tested consumer can
change execution.
\item Do not duplicate constants in documentation and consuming expressions.
Represent thresholds as one queryable policy value with validated extraction.
\item Do not use global latch, approval, or provenance state in a multi-chat bot.
\item Do not match mutation only by target head.  Bind approval to canonical full
intent and policy version.
\item Do not let Python classify moral acceptability, identity equivalence,
provenance category, agency balance, failure cause, or efficacy when those
meaning-bearing decisions belong in MeTTa.
\item Conversely, do not force hashing, authenticated transport, atomic file I/O,
or OS error normalization into MeTTa.  ``MeTTa is mind'' is an interface rule,
not substrate purism.
\item Do not overwrite belief atoms without retaining raw evidence and provenance.
\item Do not treat absence of a match as false without distinguishing missing,
malformed, stale, and explicitly false.
\item Do not retain multiple state schema versions live.  Migrate once, validate,
and delete/deactivate the old authority.
\item Do not trust comments, prompt claims, filenames, or self-map atoms as proof
of capability.  Require a probe or linked test evidence.
\item Do not add a detector without an intervention test, or an actuator without
an audit record.
\end{itemize}

\subsection{MeTTa/Python boundary contract}

For every crossing, define an input atom, output atom, error atom, version, and
idempotency key.  Python may:
\begin{itemize}
\item authenticate channel identities and obtain stable external IDs;
\item deterministically encode a MeTTa-supplied canonical typed term and compute
cryptographic hashes, without selecting its semantic identity fields;
\item perform atomic append/rename/fsync and mechanical probes;
\item normalize provider/tool mechanics into typed observations with explicit
unknown/error values, without interpreting success or policy meaning.
\end{itemize}
MeTTa should:
\begin{itemize}
\item own lifecycle state and legal transitions;
\item derive provenance categories and telemetry verdicts;
\item decide whether a mutation may reach eval;
\item validate and admit evidence, request PLN inference, and derive belief views;
\item validate continuity lineage and derive capability/self-model views.
\end{itemize}
Never pass an untyped Python string and then branch on ad hoc substrings in MeTTa.
Never let a Python exception silently imply approval, health, or success.  Return
an explicit error atom and fail closed for governance paths.

\subsection{How to test and when to stop}

Use four concentric test levels:
\begin{enumerate}
\item \textbf{Pure MeTTa truth tables:} no channels, LLMs, or files.
\item \textbf{Boundary contract tests:} fixed Python fixtures and exact atoms.
\item \textbf{Mock-loop intervention tests:} spy on provider, eval, and send to
prove that the mechanism causally changes behavior.
\item \textbf{One live-channel canary:} isolated Telegram chat, explicit logs,
no mutation of production policy until all prior gates pass.
\end{enumerate}

Stop and verify at each of these points:
\begin{itemize}
\item after defining atom shapes, before wiring writers;
\item after writers, when fresh-state and pruning invariants can be checked;
\item after detector shadow data, before enabling a consumer;
\item after consumer mock tests, before a live channel;
\item after any MeTTa runtime/library upgrade or import-path change;
\item whenever a match unexpectedly returns empty, counts remain constant, or a
prompt says a gate fired without a corresponding actuator/audit event;
\item before expanding mutation scope or turning PLN-derived ranking into policy
suppression.
\end{itemize}

Each stop should record exact commit, PeTTa/SWI-Prolog/Python versions, command,
fixtures, exit status, and observed atoms.  Preserve every discovered silent
failure as a regression test.

\section{Recommended minimum viable ingestion}

The smallest valuable release is not a ``Clarity soul.''  It is an auditable
governance kernel with:
\begin{enumerate}
\item authenticated, conversation-scoped event envelopes;
\item append-only decision provenance and a shadow agency summary;
\item exactly-once IDLE/ENGAGED/COMPLETING terminal behavior;
\item full-AST mutation proposals with expiry and an append-only continuity event;
\item immutable capability evidence plus Patham9-PLN-derived, contextual and
versioned efficacy beliefs behind an OmegaPLN-compatible adapter.
\end{enumerate}

Success should be stated behaviorally: a silent tick invokes no LLM; a terminal
event sends once; an unapproved protected mutation cannot reach eval by any skill
path; every committed self-change reconstructs from a lineage chain; and a tool's
self-reported capability status changes only from linked evidence.  If those
properties hold, the result captures ClarityOmega's strongest contributions
without inheriting its overlapping control planes and constitutional theater.

\section{Source map}

\begin{description}[style=nextline]
\item[\texttt{soul/soul\_mutation\_lock.metta}] Mutation states, approval,
expiry, commit/error reset, and lock shape.
\item[Agency-balance guard and writers] Action-tag counts, threshold,
fresh-state writer, and documented limitations.
\item[Aliveness gate and state-machine variants] Gate truth table and competing
latch representations.
\item[NACE substrate, beliefs, and capability registry] Source capability-learning
pattern, original truth-revision/courier model, dispatch filters, and fallback
semantics. NAL machinery is not recommended for ingestion.
\item[Patham9 PLN and the $\pi$PLN architecture record] Immediate inference
substrate; immutable evidence, provenance closure, explicit priors,
contextual/versioned belief views, and the OmegaPLN migration seam.
\item[\texttt{lib\_self\_continuity.metta}] Pattern-flow-network self-continuity
and chain bound.
\item[Canonical topology v08.7.2] Learning-event and lineage research schemas;
not recommended for wholesale import.
\item[\texttt{src/loop.metta}] Evidence of actual causal wiring and insertion
seams.
\item[INDEX, ADR-005, ADR-007, and investigations] Provenance, architectural
intent, stale-state failures, and detection-without-consumer failures.
\item[Hyperseed section 16 and formalization catalog] Graded self-continuity,
capability profiles, and development paths.
\item[OmegaClaw-Core at \texttt{b13b17e...}] Concrete target file/interface map.
\end{description}

\end{document}
